---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
```

# reptiledbr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/reptiledbr)](https://CRAN.R-project.org/package=reptiledbr)
[![R-CMD-check](https://github.com/PaulESantos/reptiledbr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/reptiledbr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PaulESantos/reptiledbr/graph/badge.svg)](https://app.codecov.io/gh/PaulESantos/reptiledbr)
<!-- badges: end -->

## Overview

**`reptiledbr`** is an R package that provides programmatic access to data from [The Reptile Database](http://www.reptile-database.org), a comprehensive and curated source of taxonomic information on all living reptiles. This includes snakes, lizards, turtles, tuataras, amphisbaenians, and crocodiles—over 10,000 species and more than 2,800 subspecies.

The goal of `reptiledbr` is to facilitate access to reptile taxonomy, nomenclature, distribution, and associated scientific literature in a reproducible and efficient way for researchers, ecologists, educators, and biodiversity data users.


## Installation

You can install the development version of `reptiledbr` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("PaulESantos/reptiledbr")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(reptiledbr)
## basic example code


reptiledb <- reptiledbr::get_reptiledb_data("Anolis carolinensis", quiet = FALSE)

reptiledb |> 
  reptiledbr::format_all_attributes()
```

## Features

- Query species by scientific name or higher taxonomy
- Retrieve synonyms, type localities, and distributions
- Access bibliographic references for species descriptions
- Tools to explore phylogenetic classification (based on Zheng & Wiens, 2016)
- Fully integrated with the R environment for analysis and visualization

---

## Data Source

The package uses data from The Reptile Database, an open and community-driven resource curated by herpetologists and volunteers around the world. It focuses on taxonomic data including:

- Scientific names and synonyms

- Type specimens and distribution

- Original literature references

- The database was initiated by Peter Uetz in 1995 and is currently curated by a volunteer team. For more information, visit: www.reptile-database.org

> Note: This package is not officially affiliated with The Reptile Database. It provides an independent R interface to facilitate access to public data.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
library(viridis)  # For an accessible and attractive color palette
library(hrbrthemes)  # For modern themes
library(scales)  # For formatting labels

# Create the tibble with the data
species_data <- tribble(
  ~Year, ~Amphisbaenia, ~Sauria, ~Serpentes, ~Testudines, ~Crocodylia, ~Rhynchocephalia, ~Total,
  "Feb 2008", 168, 5079, 3149, 313, 23, 2, 8734,
  "Jan 2011", 181, 5461, 3315, 317, 24, 2, 9300,
  "Feb 2012", 181, 5634, 3378, 327, 25, 2, 9547,
  "Feb 2013", 184, 5796, 3432, 328, 25, 1, 9766,
  "Aug 2014", 188, 5987, 3496, 341, 25, 1, 10038,
  "Aug 2015", 193, 6145, 3567, 341, 25, 1, 10272,
  "Aug 2016", 196, 6263, 3619, 346, 25, 1, 10450,
  "Oct 2017", 193, 6399, 3672, 350, 24, 1, 10639,
  "Jul 2018", 196, 6512, 3709, 351, 24, 1, 10793,
  "Aug 2019", 195, 6687, 3789, 353, 25, 1, 11050,
  "Aug 2020", 201, 6905, 3848, 360, 26, 1, 11341,
  "Nov 2021", 202, 7144, 3956, 360, 27, 1, 11690,
  "Dec 2022", 201, 7310, 4038, 363, 27, 1, 11940,
  "Oct 2023", 202, 7396, 4073, 361, 27, 1, 12060,
  "Aug 2024", 203, 7724, 4145, 366, 27, 1, 12263,
  "Jan 2025", 203, 7814, 4177, 367, 27, 1, 12386
)

# Transform data to long format for visualization
species_long <- species_data %>% 
  pivot_longer(cols = -Year, 
               names_to = "Group",
               values_to = "Species") %>% 
  filter(Group != "Total") %>%
  # Extract year for better labeling
  mutate(
    Year_num = as.numeric(substr(Year, nchar(Year)-3, nchar(Year))),
    # Reorder groups in a more logical presentation
    Group = factor(Group, 
                   levels = c("Sauria", "Serpentes", "Testudines", 
                              "Amphisbaenia", "Crocodylia", "Rhynchocephalia"))
  )

# Create a custom color palette and more descriptive names
group_colors <- viridis_pal(option = "turbo")(6)
group_labels <- c("Sauria (Lizards)", "Serpentes (Snakes)", "Testudines (Turtles)", 
                 "Amphisbaenia (Worm Lizards)", "Crocodylia (Crocodilians)", 
                 "Rhynchocephalia (Tuataras)")
names(group_labels) <- levels(species_long$Group)

```


```{r echo=FALSE, warning=FALSE, message=FALSE}
p <- ggplot(species_long, 
       aes(x = Year, y = Species, color = Group, group = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = group_colors, labels = group_labels) +
  labs(
    title = "Evolution of Reptile Species (2008–2025)",
    subtitle = "Data from The Reptile Database",
    x = "",
    y = "Number of Species",
    color = "Taxonomic Group",
    caption = "Source: The Reptile Database | Visualization: {reptiledbr}"
  ) +
  theme_ipsum_rc(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "grey80", fill = NA),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, margin = margin(b = 20)),
    plot.caption = element_text(size = 9, color = "grey40"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  )

# Create a separate plot for each group with appropriate scales
p_facet <- p + 
  facet_wrap(~Group, scales = "free_y", ncol = 2) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )

# Create a trend plot (only for the main groups)
p_trend <- species_long %>%
  filter(Group %in% c("Sauria", "Serpentes", "Testudines")) %>%
  ggplot(aes(x = Year, y = Species, color = Group, group = Group)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = group_colors[1:3], labels = group_labels[1:3]) +
  labs(
    title = "Trends in Major Reptile Groups",
    x = "",
    y = "Number of Species",
    color = "Taxonomic Group"
  ) +
  theme_ipsum_rc(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

# Display the plots
print(p_facet)
print(p_trend)
```


## Citing the Data

Please cite the original Reptile Database if you use this package in published work:

Uetz, P., Freed, P., & Hošek, J. (eds.) (2021). The Reptile Database. Retrieved from http://www.reptile-database.org

## Contributing

Contributions, bug reports, and feature requests are welcome! Please use the issue tracker to report problems or suggest improvements.

## License

This package is free and open source software, licensed under the MIT License. See the LICENSE file for more details.

## Acknowledgments

Special thanks to the editors and contributors of The Reptile Database for their dedication to herpetological taxonomy and for maintaining an open-access scientific resource for the global community.


