---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# reptiledbr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/reptiledbr)](https://CRAN.R-project.org/package=reptiledbr)
[![R-CMD-check](https://github.com/PaulESantos/reptiledbr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/reptiledbr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PaulESantos/reptiledbr/graph/badge.svg)](https://app.codecov.io/gh/PaulESantos/reptiledbr)
<!-- badges: end -->

## Overview

**`reptiledbr`** is an R package that provides programmatic access to data from [The Reptile Database](http://www.reptile-database.org), a comprehensive and curated source of taxonomic information on all living reptiles. This includes snakes, lizards, turtles, tuataras, amphisbaenians, and crocodiles—over 10,000 species and more than 2,800 subspecies.

The goal of `reptiledbr` is to facilitate access to reptile taxonomy, nomenclature, distribution, and associated scientific literature in a reproducible and efficient way for researchers, ecologists, educators, and biodiversity data users.


## Installation

You can install the development version of `reptiledbr` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("PaulESantos/reptiledbr")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(reptiledbr)
## basic example code


reptiledb <- reptiledbr::get_reptiledb_data("Anolis carolinensis", quiet = FALSE)

reptiledb |> 
  reptiledbr::format_all_attributes()
```

## Features

- Query species by scientific name or higher taxonomy
- Retrieve synonyms, type localities, and distributions
- Access bibliographic references for species descriptions
- Tools to explore phylogenetic classification (based on Zheng & Wiens, 2016)
- Fully integrated with the R environment for analysis and visualization

---

## Data Source

The package uses data from The Reptile Database, an open and community-driven resource curated by herpetologists and volunteers around the world. It focuses on taxonomic data including:

- Scientific names and synonyms

- Type specimens and distribution

- Original literature references

## Data Source

The package uses data from The Reptile Database, an open and community-driven resource curated by herpetologists and volunteers around the world. It focuses on taxonomic data including:

- Scientific names and synonyms

- Type specimens and distribution

- Original literature references

- The database was initiated by Peter Uetz in 1995 and is currently curated by a volunteer team. For more information, visit: www.reptile-database.org

> Note: This package is not officially affiliated with The Reptile Database. It provides an independent R interface to facilitate access to public data.

## Citing the Data

Please cite the original Reptile Database if you use this package in published work:

Uetz, P., Freed, P., & Hošek, J. (eds.) (2021). The Reptile Database. Retrieved from http://www.reptile-database.org

## Contributing

Contributions, bug reports, and feature requests are welcome! Please use the issue tracker to report problems or suggest improvements.

## License

This package is free and open source software, licensed under the MIT License. See the LICENSE file for more details.

## Acknowledgments

Special thanks to the editors and contributors of The Reptile Database for their dedication to herpetological taxonomy and for maintaining an open-access scientific resource for the global community.


